%! Author = clem
%! Date = 23.11.20

\input{settings} % load settings

\usepackage[latin9]{luainputenc}
\usepackage{color}
\usepackage{array}
\usepackage{float}
\usepackage{textcomp}
\usepackage{amsthm}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}
\floatstyle{ruled}
\newfloat{algorithm}{tbp}{loa}
\providecommand{\algorithmname}{Algorithm}
\floatname{algorithm}{\protect\algorithmname}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Textclass specific LaTeX commands.
\newenvironment{lyxcode}
	{\par\begin{list}{}{
		\setlength{\rightmargin}{\leftmargin}
		\setlength{\listparindent}{0pt}% needed for AMS classes
		\raggedright
		\setlength{\itemsep}{0pt}
		\setlength{\parsep}{0pt}
		\normalfont\ttfamily}%
	 \item[]}
	{\end{list}}
\theoremstyle{plain}
    \ifx\thechapter\undefined
      \newtheorem{lem}{\protect\lemmaname}
    \else
      \newtheorem{lem}{\protect\lemmaname}[chapter]
    \fi

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\definecolor{gray}{rgb}{0.5, 0.5, 0.5}
\definecolor{keyword}{RGB}{243, 60, 114}
\definecolor{comment}{RGB}{0, 200, 101}
\definecolor{background}{rgb}{0.9, 0.9, 0.9}

\usepackage{xparse}
\usepackage{algorithm,algpseudocode}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}

\makeatother

\lstset{keywordstyle={\bfseries \color{keyword}},
commentstyle={\color{comment}},
caption={Search function for exact matches for a query of size 0 < m < k $^{[1]}$},
captionpos=b,
backgroundcolor={\color{background}},
basicstyle={\ttfamily\small},
language={C++},
numbers=none,
captionpos=b,
tabsize=2}
\providecommand{\lemmaname}{Lemma}
\renewcommand{\lstlistingname}{Listing}

\begin{document}

% thesis details
\title{Implementation and Performance Evaluation of a kmer Index intended for Exact String Matching}
\newcommand{\Year}{2020}
\date{TODO\textsuperscript{st} December \Year2020}
\newcommand{\thesisKind}{Bachelor}
\newcommand{\degree}{Bachelor~of~Science}
\newcommand{\versionnumber}{1.0}

% author details
\author{Clemens~Cords}
\newcommand{\matrikelno}{4872639}
\newcommand{\email}{clemens.cords@fu-berlin.de}

% supervisor details
\newcommand{\supervisor}{Prof. Dr. Knut Reinert}
\newcommand{\supervisorUniversity}{Freie Universit\"at Berlin}
\newcommand{\supervisorDepartment}{Dept. of Computer Science and Mathematics}
\newcommand{\supervisorAG}{Algorithmische Bioinformatik}
\newcommand{\supervisorCountry}{Germany}

% Advisor 1 details
\newcommand{\fstAdvisor}{Enrico Seiler}
\newcommand{\fstAdvisorsUniversity}{Freie Universit\"at Berlin}
\newcommand{\fstAdvisorsDepartment}{Dept. of Computer Science and Mathematics}
\newcommand{\fstAdvisorsAG}{Algorithmische Bioinformatik}
\newcommand{\fstAdvisorsCountry}{Germany}

% Advisor 2 details
\newcommand{\sndAdvisor}{Prof. Dr. Knut Reinert}
\newcommand{\sndAdvisorsUniversity}{Freie Universit\"at Berlin} % or company
%\newcommand{\sndAdvisorsDepartment}{Dept. of Computer Science and Mathematics}
%\newcommand{\sndAdvisorsAG}{Algorithmische Bioinformatik}
\newcommand{\sndAdvisorsCountry}{Germany}

\include{titlepage}

% Abstract in english
\begin{abstractEN}
As demand for large-scale genomic sequence matching rises, as does
demand for the ability to efficiently compute the positions or number of occurrences
of a nucleotide sequence of length k called a kmer. A kmer-index optimized for this
purposes was implemented and it's performance evaluated. The implementation
is capable of searching kmers of arbitrary length specified at runtime,
utilizes parallelization for faster construction and it's internal
structure can be further customized at compile time to achieve greater
performance for certain query sizes. The nature of the internal structure
and search functions are explained and their performant nature demonstrated
through benchmarks. Results indicate that for the purpose of finding
positions or number of occurrences for queries of
length $k\in\{3,4,...,30\}$ compared to seqan3s fm\_index the kmer-index
performs up to 60\% faster and is thus recommended for use. For arbitrarily
long queries the kmer-index was shown to exhibit speedup or slowdown
of $\pm5$\% depending on text length and is therefore in it's current
iteration applicable if not necessarily better suited for this purpose.
\end{abstractEN}
\vfill

% abstract in german
\begin{abstractDE}
Mit steigender Nachfrage f\"ur gro\ss fl\"achige genomisches Sequence Matching steigt
auch die Nachfrage Positionen f\"ur kurze genomische Sequenzen oder kmer performant
in einem Text zu finden. Zu diesem Zweck wurde ein sogenannter kmer-Index implementiert
und dessen Performance evaluiert. Die Implementierung ist f\"ahig kmer von willk\"urlicher
L\"ange zu finden, benutzt Parallelisierung f\"ur schneller Konstruktion und die interne
Struktur des Indices kann weiterhin personalisiert werden um bessere Performance
f\"ur bestimtte Query L\"angen zu erhalten. Die Natur der Personalisierung und die daraus 
resultierende Performance wurde mit Benchmarks demonstriert. Ergebnisse legen nah, 
dass f\"ur kmer der L\"ange $k\in\{3,4,...,30\}$ im Vergleich zu seqan3s fm\_index der 
kmer-index bis zu 60\% schneller ist. F\"ur kmer beliebiger L\"ange zeigen Ergebnisse, 
dass ein Speedup von $\pm5$\% je nach Textl\"ange erreicht wird. Der Einsatz der kmer-Index
Implementierung wird daher f\"ur kurze kmer empfohlen w\"ahrend f\"ur l\"angere kmer der Index 
je nach Text L\"ange einsetzbar aber nicht unbedingt besser geeignet ist.
\end{abstractDE}
\vfill

%%% TOC %%%
\tableofcontents{}
\setcounter{page}{1}

% 01: Introduction and motivation
\include{content/01_introduction}

% 02: Implementation
\include{content/02_implementation}

% 03: Methods
\include{content/03_measuring_performance}

% 04: Outlook
\include{content/04_outlook_and_conclusion}

% 05: Addendum
\newpage{}
\include{content/05_addendum}

\pagebreak{}

% Bibliography
\bibliography{references}
\bibliographystyle{ieeetr}

\end{document}